---
title: "Class 7: Machine Learning 1"
author: Christopher Leone | A16731724
format: html
toc: true
title-block-banner: "navy"
---

> Today we will explore unsupervised machine learning methods starting with clustering and dimensionality reduction.

## Clustering

To start, let's make up some data to cluster where we know what the answer should be. The `rnorm()` function will help us here.

```{r}
# This will create a histogram displaying 10000 randomly generated numbers.
hist(rnorm(10000))
```

And here, we create a plot with two clusters of data located at the arbitrary x values of "-3" and "3", with y values reversed.

```{r}
# This binds the two rnorm() groups inside an object to plot
val <- c(rnorm(30, 3), rnorm(30, -3))
x <- cbind(x=val, y=rev(val))
plot(x)
```

### K-means Clustering

The main function in "base R" for K-means clustering is called `kmeans()`:

```{r}
# Only 'x' and 'centers' are required args for `kmeans()`
k <- kmeans(x, 2)
k
```

The `kmeans()` function returns a "list" with 9 components. You can see the named components with the `attributes()` function.

```{r}
# This lists the 9 attributes.
attributes(k)
```

> (Q) How many points are in each cluster?

```{r}
# Size gives the size of each cluster.
k$size
```
> (Q) Cluster assignment/membership vector?

```{r}
# This calls the vector for each cluster assignment.
k$cluster
```

> (Q) Cluster centers?

```{r}
# This returns the coordinates of the cluster centers.
k$centers
```

> (Q) Make a plot of our `kmeans()` results showing cluster assignment with different colors for each group of points and centers in blue.

```{r}
# Col can also be a number, so we can set it equal to k$cluster.
plot(x, col=(k$cluster))
points(k$centers, col="blue", pch=17, cex=2)
```

> (Q) Run `kmeans()` again on `x` and this time cluster into four groups and plot the same result figure as above.

```{r}
k2 <- kmeans(x,4)
plot(x, col=(k2$cluster))
points(k2$centers, col="blue", pch=17, cex=2)
```

> **Key Point**: K-means clustering is very popular but can be misused. One big limitation is that it can impose a clustering pattern on your data even if clear natural grouping doesn't exist - i.e. it does what you tell it to do in terms of `centers`.

### Hierarchical Clustering

The main function in "base R" for hierarchical clustering is `hclust()`.

You can't just pass our dataset into `hclust()`, more work is required, but it is more flexible than `kmeans()`.

The results of `hclust()` don't have a useful `print()` method but do have a special `plot()` method.

```{r}
# hclust() only has one required input, "d".
d <- dist(x)
h <- hclust(d)
plot(h)
```
To get our main cluster assignment (membership vector), we need to "cut" the tree at the big "goalposts"...

> (Q) How many large clusters are there?

```{r}
# This will return our two original clusters!
grps <- cutree(h, h=8)
table(grps)
```
Hierarchical clustering is distinct in that the dendrogram can reveal the potential grouping in your data, unlike `kmeans()`.

## Dimensionality | Principal Component Analysis

PCA is a common and highly useful dimensionality reduction technique used in many fields - particularly in bioinformatics.

Here we will analyze some data from the UK on food consumption.

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
head(x)
```

> (Q1) How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?

```{r}
dim(x)
```

But what if we don't want X to be counted as a column of data? 

> (Q2) Here is the preferred solution, because it is does not continually delete column names each time the code is ran:

```{r}
# This is a non-destructive method
x <- read.csv(url, row.names=1)
head(x)
```

...and the new dimensions:

```{r}
dim(x)
```

### Spotting Major Differences

First, let's paste the original plot:

```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```

> (Q3) Changing what optional argument in the above barplot() function results in the following plot?

Let's make this plot different by changing one parameter:

```{r}
# "beside" makes a side-by-side barplot, we change it to "F".
barplot(as.matrix(x), beside=F, col=rainbow(nrow(x)))
```

This plot is still not very helpful, so lets try a different conventional method — a "pairs" plot.

```{r}
# But, this is still quite complicated...
pairs(x, col=rainbow(nrow(x)), pch=16)
```

### PCA to the Rescue!

The main function in "base R" for PCA is `prcomp()`:

```{r}
# But, this function wants this data in a specific way.
# So, lets transpose our matrix:
pca <- prcomp(t(x))
summary(pca)
```

The `prcomp()` function returns a list object of our results with 5 components, according to `attributes()`.

The two main results in here are `pca$x` and `pca$rotation`. The first of these contains the scores of the data on the new PC axis, which we use to make our PCA plot.

> (Q7) Complete the code below to generate a plot of PC1 vs PC2. 

```{r}
# Calling the gg libraries...
library(ggplot2)
library(ggrepel)

#...then making a ggplot of PC1 vs PC2.
ggplot(pca$x) +
  aes(PC1, PC2, label=rownames(pca$x)) +
  geom_point() +
  geom_text_repel()
```

> What we can gather from this is that Wales, England, and Scotland all fall neatly along the PC1 line, but N. Ireland is significantly different. Let's see why:

We can also use `pca$rotation` to determine what in the data is leading to the variation we see in our plot.

```{r}
# Calling the library...
library(ggplot2)

# And creating the plot.
ggplot(pca$rotation) +
  aes(PC1, rownames(pca$rotation)) +
  geom_col()
```

By reading both PCA plots, we can determine that:

- N. Ireland is different because they consume considerably more fresh potatoes and soft drinks.
- The rest consume considerably more alcoholic drinks and fresh fruit.

Additionally, we can see this correlation with biplots. (`biplot()` function)

```{r}
# This is found in "base R".
biplot(pca)
```

