---
title: "Class 9 | Halloween Mini-Project"
author: Christopher Leone | A16731724
format: html
toc: true
---

## Background

Today we will take a step back to some data we can taste and explore—we will examine the correlation structure and principal components of some Halloween candy.

```{r}
# Let's read the data into our project:
candy <- read.csv("candy-data.csv", row.names=1)
head(candy)
```

> (Q1) How many different candy types are in this dataset?

```{r}
# Shows that we have 85 entries.
nrow(candy)
```

> (Q2) How many fruity candy types are in the dataset?

```{r}
# 38 fruity candies!
sum(candy$fruity)
```

## What is your favorite candy?

> (Q3) What is your favorite candy in the dataset and what is it’s winpercent value?

My favorite Halloween candy are the Nestle Crunch candies. How popular is it?

```{r}
candy["Nestle Crunch",]$winpercent
```

> (Q4) What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat",]$winpercent
```

> (Q5) What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```

A helpful function we can use to get a summary of a dataset is the `skim()` function from the **skimr** package, which we downloaded and can call to examine our `candy` dataset.

```{r}
library(skimr)
skim(candy)
```

> (Q6) Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The **winpercent** column appears to be on a different scale because it is the only column without values consistently below 1.

> (Q7) What do you think a zero and one represent for the candy$chocolate column?

Each "zero" represents a "No/FALSE", while each "one" represents a "Yes/TRUE", such as in binary code.

> (Q8) Plot a histogram of winpercent values.

```{r}
# Let's just generate a simple histogram here
hist(candy$winpercent)
```

> (Q9) Is the distribution of winpercent values symmetrical?

No, the distribution is swayed to the left, closer to lower `candy$winpercent` values.

> (Q10) Is the center of the distribution above or below 50%?

As alluded to in **Q8**, the center of the distribution is below 50%.

> (Q11) On average is chocolate candy higher or lower ranked than fruit candy?

Let's check the values for each:

```{r}
# Here is the mean value for chocolate candy:
mean(candy$winpercent[as.logical(candy$chocolate)])

# And here is the mean for fruity candy:
mean(candy$winpercent[as.logical(candy$fruity)])
```

We can see that chocolate candy has **~16%** higher `winpercent` values than fruity candy.

> (Q12) Is this difference statistically significant?

```{r}
# We can use a T-test to find the p-value and significance:
t.test(x=candy$winpercent[as.logical(candy$chocolate)],y=candy$winpercent[as.logical(candy$fruity)])
```

We get a very small p-value of **2.871e-08**, which indicates a very statistically significant difference between the `winpercent` of chocolate and fruity candies.

## Overall Candy Rankings

We can make this easier by ordering the data by `winpercent`, which would allow us to see the most/least liked candies.

> (Q13) What are the five least liked candy types in this set?

There are two related functions that can help here, one is the classic `sort()` function and the `order()` function. But, here let's play with the `dplyr` package.

```{r}
# Let's experiment with the dplyr package!
library(dplyr)
candy %>% arrange(winpercent) %>% head(5)
```

The 5 least liked candies in the dataset are "Nik L Nip", "Boston Baked Beans", "Chiclets", "Super Bubble", and "Jawbusters".

> (Q14) What are the top 5 all time favorite candy types out of this set?

To solve this in a similar manner, we can use the `tail()` function.

```{r}
# Let's experiment with the dplyr package!
library(dplyr)
candy %>% arrange(winpercent) %>% tail(5)
```

According to the data, the 5 most liked candies are "Reese's Peanut Butter Cups", "Reese's Miniatures", "Twix", "Kit Kat", and "Snickers".

### Making a Barplot

Let's make an intricate barplot to analyze the overall rankings of the candy. We will start simple and build from there.

> (Q15) Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)
ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col()
```

> (Q16) This is quite ugly, use the reorder() function to get the bars sorted by `winpercent`.

We can use the `reorder()` function within the `aes()` part of creating a **ggplot** by giving it a variable to sort by.

```{r}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy), winpercent)) + 
  geom_col()
```

I can make it even prettier by adding colors to the bars. Let's color by candy type *(the code is borrowed from the Class Lab worksheet)*, and reprint the barplot.

```{r}
# Here we define our color categories,
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "maroon"

# And here we reprint our barplot.
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 

# Why not save it?
ggsave("my_ugly_barplot.png", width=6, height=9)
```

Using our new colored plot...

> (Q17) What is the worst ranked chocolate candy?

The lowest ranked chocolate candy is "Sixlets".

> (Q18) What is the best ranked fruity candy?

The highest ranked fruity candy is "Starburst".

## Taking a look at `pricepercent`

Let's compare `pricepercent` and `winpercent` to see which candy will give us the greatest bang for our buck. We want lower `pricepercent` and higher `winpercent`.

```{r}
library(ggrepel)

# How about a plot of price vs win?
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

> (Q19) Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
ord <- order(candy$winpercent, decreasing = TRUE)
head(candy[ord,c(11,12)], n=5)
```


We can see that the best candies with regards to both `pricepercent` and `winpercent` are **Reese's Miniatures**.

> (Q20) What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head(candy[ord,c(11,12)], n=5)
```

The top 5 most expensive candy types are **Nik L Nip, Nestle Smarties, Ring Pops, Hershey's Krackel, and Hershey's Milk Chocolate**. Out of these 5, the least popular is **Nik L Nip**, which is also the most expensive! Not a great choice...

## Exploring the Correlation Structure

Let's check the correlation between different candy characteristics to see what we can find.

```{r}
# We have to download the `corrplot` library first
c <- cor(candy)
corrplot::corrplot(c)
```

> (Q22) Examining this plot what two variables are anti-correlated (i.e. have minus values)?

The strongest anti-correlation on this plot is between **chocolate** and **fruity**, with a **-0.74** correlation.

> (Q23) Similarly, what two variables are most positively correlated?

Outside of the 1.00 correlation between the same variable, the greatest positive correlation is between **chocolate** and **winpercent**, with a **+0.64** correlation.

## Principal Component Analysis

We need to be sure to scale our data, because we recall that `winpercent` is on a different scale than the rest of the data. 

```{r}
# Our first two PCAs only account for 47%...
pca <- prcomp(candy, scale=T)
summary(pca)
```

First, lets make a "PCA plot":

```{r}
ggplot(pca$x) +
  aes(PC1, PC2, label=rownames(pca$x)) +
  geom_point(col=my_cols) +
  geom_text_repel(max.overlaps = 6, col=my_cols) +
  theme_bw()
```

The second main PCA result is in the `pca$rotation` object; we can plot this to generate a so-called "loadings" plot.

```{r}
ggplot(pca$rotation) +
  aes(PC1, reorder(rownames(pca$rotation), PC1), fill=PC1) +
  geom_col()
```

> (Q24) What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

The strongest positively correlated variables with PC1 are **fruity, pluribus, and hard**. This makes sense, because many fruity candies come in packs with more than one candy, and are often also hard candies.



