---
title: "Class 18: Pertussis Mini-Project"
author: "Christopher Leone | A16731724"
format: html
toc: true
---

## Background:

Pertussis, commonly known as whooping cough, is a common lung infection caused by the bacterium *Bordetella pertussis*. 

The CDC tracks cases of pertussis in the US: https://tinyurl.com/pertussiscdc 

## Examining Cases of Pertussis by Year:

First, we have to scrape the data in from the website using **DATAPASTA**.

```{r, include=FALSE}
cdc <- data.frame(year = c(1922L,1923L,1924L,1925L,1926L,1927L,1928L,
                           1929L,1930L,1931L,1932L,1933L,1934L,1935L,
                           1936L,1937L,1938L,1939L,1940L,1941L,1942L,
                           1943L,1944L,1945L,1946L,1947L,1948L,1949L,
                           1950L,1951L,1952L,1953L,1954L,1955L,1956L,
                           1957L,1958L,1959L,1960L,1961L,1962L,1963L,
                           1964L,1965L,1966L,1967L,1968L,1969L,1970L,
                           1971L,1972L,1973L,1974L,1975L,1976L,1977L,
                           1978L,1979L,1980L,1981L,1982L,1983L,1984L,
                           1985L,1986L,1987L,1988L,1989L,1990L,1991L,
                           1992L,1993L,1994L,1995L,1996L,1997L,1998L,
                           1999L,2000L,2001L,2002L,2003L,2004L,2005L,
                           2006L,2007L,2008L,2009L,2010L,2011L,2012L,
                           2013L,2014L,2015L,2016L,2017L,2018L,2019L,
                           2020L,2021L,2022L,2023L,2024L),
             pertussis = c(107473,164191,165418,152003,202210,181411,
                           161799,197371,166914,172559,215343,179135,
                           265269,180518,147237,214652,227319,103188,
                           183866,222202,191383,191890,109873,133792,
                           109860,156517,74715,69479,120718,68687,
                           45030,37129,60886,62786,31732,28295,
                           32148,40005,14809,11468,17749,17135,
                           13005,6799,7717,9718,4810,3285,4249,
                           3036,3287,1759,2402,1738,1010,2177,2063,
                           1623,1730,1248,1895,2463,2276,3589,
                           4195,2823,3450,4157,4570,2719,4083,6586,
                           4617,5137,7796,6564,7405,7298,7867,
                           7580,9771,11647,25827,25616,15632,10454,
                           13278,16858,27550,18719,48277,28639,
                           32971,20762,17972,18975,15609,18617,6124,
                           2116,3044,7063,35493)
)
```

> (Q1): Make a plot of Pertussis cases per year using ggplot.

```{r}
library(ggplot2)
cases <- ggplot(cdc) +
            aes(year, pertussis) +
            geom_point() + 
            geom_line()
```

> (Q2): Add some key time points in our graph. These include the years when wP was developed (1946) and the switch to aP (1996).

We can use `geom_vline()` to create vertical lines on our graph to denote these checkpoints.

```{r}
cases +
  geom_vline(xintercept=1946, col="blue") +
  geom_vline(xintercept=1996, col="red") +
  geom_vline(xintercept=2020, col="purple")
```

Mounting evidence suggests that the newer **aP** (acellular vaccine) is less effective over the long term than the original **wP** vaccine. In other words, vaccine protection wanes more rapidly with **aP**.

## Enter the CMI-PB Project:

The CMI-PD project (Computational Models of Immunity - Pertussis Boost) aims to investigate how the immune system responds differently in individuals with the **aP** or **wP** vaccines. The goal is to be able to predict when the immunity fades so we know the true efficacy of each vaccine.

CMI-PB makes all of their collected data publicly available and they store it in a database composed of different tables. Herem, we will access a few of these.

We can use the `jsonlite` package to read this data.

```{r}
library(jsonlite)
subject <- read_json("https://www.cmi-pb.org/api/v5_1/subject",
                     simplifyVector=TRUE)
head(subject)
```

> (Q): How many subjects are there?

According to `nrow()`, there are 172 participants in this dataset.

```{r}
nrow(subject)
```

> (Q): How many "aP" and "wP" individuals are there?

Using `table()`, we see that there are 87 aP-vaccinated individuals, and 85 wP-vaccinated individuals.

```{r}
table(subject$infancy_vac)
```

> (Q): How many M/F subjects are in the dataset?

Using `table()`, we see that there are 112 female individuals, and 60 male individuals.

```{r}
table(subject$biological_sex)
```

> (Q): How many M/F of each race are there in the dataset?

This dataset is dominated by White Females, followed by Asian Females and White Males.

```{r}
table(subject$race, subject$biological_sex)
```

> (Q): Is this representative of the US Demographics?

No way! It looks like UC San Diego's population...

Let's go read some more data from CMI-PB:

```{r}
specimen <- read_json("https://www.cmi-pb.org/api/v5_1/specimen",
                      simplifyVector = TRUE)
ab_data <- read_json("http://cmi-pb.org/api/v5_1/plasma_ab_titer",
                      simplifyVector = TRUE)
head(specimen)
head(ab_data)
```

Let's join our `subject` and `specimen` tables to get all of our information together using a **dplyr** function.

```{r}
library(dplyr)
meta <- inner_join(subject, specimen)
head(meta)
```

We can also join our `meta` table with our `ab_data`, so we can have one large dataset.

```{r}
data <- inner_join(ab_data, meta)
head(data)
```

> (Q): How many antigen types are measured? 
> (Q): How many total antigens are there in the data?

Out of our 61956 antigens measured in the dataset, there are 15 studied antigens.

```{r}
table(data$antigen)
nrow(data)
```

> (Q): Make a boxplot of antigen levels across the whole dataset.

```{r}
ggplot(data) +
  aes(MFI, antigen) +
  geom_boxplot()
```

> (Q): Are there obvious differences between **aP** and **wP** individuals?

```{r}
ggplot(data) +
  aes(MFI, antigen, col=infancy_vac) +
  geom_boxplot()
```

## Focus on IgG

Here, we can use a filter to focus on the most abundant immune antigen -> IgG.

Let's make a similar boxplot of antigens.

```{r}
## Defining IgG
igg <- data |> filter(isotype == "IgG")

## Making the plot
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac) +
  geom_boxplot() +
  facet_wrap(~visit)
```

Let's zoom in even further on PT (pertussis toxin) levels over time, within the **2021 dataset**.

```{r}
# Filtering the data
pt <- data |> filter(isotype=="IgG",
                     antigen=="PT",
                     dataset=="2021_dataset")
# Creating a plot
ggplot(pt) +
  aes(actual_day_relative_to_boost,
      MFI_normalised,
      col=infancy_vac,
      group=subject_id) +
  geom_point() +
  geom_line() +
  theme_bw() +
  geom_vline(xintercept=14, col="black")
```



