---
title: "Class 17: Analyzing Sequencing Data"
author: "Christopher Leone | A16731724"
format: html
toc: true
---

## Downstream Analysis

After downloading the `tximport` package, we can directly and easily read Kallisto results.

```{r}
library(tximport)

# Setup the folder and filenames to read
folders <- dir(pattern="SRR21568*")
samples <- sub("_quant", "", folders)
files <- file.path( folders, "abundance.h5" )
names(files) <- samples

txi.kallisto <- tximport(files, type = "kallisto", txOut = TRUE)

# Taking a look:
head(txi.kallisto$counts)
```

> (Q): How many transcripts do we have for each sample?

The results are shown by `colSums()`.

```{r}
colSums(txi.kallisto$counts)
```

> (Q): How many transcripts are in at least one sample?

There were 94,561 transcripts that were present in at least one sample.

```{r}
sum(rowSums(txi.kallisto$counts)>0)
```

## Principal Component Analysis

Here is some code to filter the results:

```{r}
to.keep <- rowSums(txi.kallisto$counts) > 0
kset.nonzero <- txi.kallisto$counts[to.keep,]
```

```{r}
keep2 <- apply(kset.nonzero,1,sd)>0
x <- kset.nonzero[keep2,]
```


...Before moving onto PCA.

```{r}
pca <- prcomp(t(x), scale=TRUE)

summary(pca)
```

> (Q): Make PCA Plots of PC1 vs PC2.

First, let's load in some additonal code. We are loading in the metadata as well as creating teh dataframe so we can make a `ggplot` rather than a `base R` plot.

```{r}
# Make metadata object for the samples
colData <- data.frame(condition = factor(rep(c("control", "treatment"), each = 2)))
rownames(colData) <- colnames(txi.kallisto$counts)

# Make the data.frame for ggplot 
y <- as.data.frame(pca$x)
y$Condition <- as.factor(colData$condition)
```

Now, here' PC1 vs. PC2:

```{r}
library(ggplot2)
library(ggrepel)

ggplot(y) +
  aes(PC1, PC2, col=Condition) +
  geom_point() +
  geom_text_repel(label=rownames(y)) +
  theme_bw()
```

> (Q): Make PCA Plots of PC1 vs PC3.

The process is the same as before:

```{r}
ggplot(y) +
  aes(PC1, PC3, col=Condition) +
  geom_point() +
  geom_text_repel(label=rownames(y)) +
  theme_bw()
```

> (Q): Make PCA Plots of PC2 vs PC3.

The process is the same as before, one final time:

```{r}
ggplot(y) +
  aes(PC2, PC3, col=Condition) +
  geom_point() +
  geom_text_repel(label=rownames(y)) +
  theme_bw()
```