---
title: 'Lab 6 Homework: Question 6'
author: "Christopher Leone | A16731724"
date: "2025-04-20"
output: pdf_document
toc: true
---

# Preceding Code Chunk

## Here is the code chunk preceding the main code to be edited in Q6.


```{r message=FALSE, warning=FALSE}
# This calls upon bio3d
library(bio3d)

s1 <- read.pdb("4AKE") # kinase with drug
s2 <- read.pdb("1AKE") # kinase no drug
s3 <- read.pdb("1E4Y") # kinase with drug

# This trims a predefined protein into various chains into a smaller PDB object.
s1.chainA <- trim.pdb(s1, chain="A", elety="CA")
s2.chainA <- trim.pdb(s2, chain="A", elety="CA")
s3.chainA <- trim.pdb(s1, chain="A", elety="CA")

# This gathers the various B-factor values to be plotted
s1.b <- s1.chainA$atom$b
s2.b <- s2.chainA$atom$b
s3.b <- s3.chainA$atom$b

# The B-factor plots will not be printed here for sake of space.
```


# Main Assignment: Question 6

```{r message=FALSE, warning=FALSE}
hc <- hclust( dist( rbind(s1.b, s2.b, s3.b) ) )
plot(hc)
```


## (Q6): How would you generalize the original code above to work with any set of input protein structures?

### 1) If we want to generalize the original code above, we have to rely on creating a `function()`. But first, I want to define similar variables to `s1.b, s2.b, s3.b` to add to the object pool.


```{r message=FALSE, warning=FALSE}
# This calls upon bio3d
library(bio3d)

s4 <- read.pdb("1O6K") # PKB Kinase
s5 <- read.pdb("2BDJ") # Src Kinase (inhibited)
s6 <- read.pdb("6F3F") # Src Kinase (not inhibited)

# This trims a predefined protein into various chains into a smaller PDB object.
s4.chainA <- trim.pdb(s4, chain="A", elety="CA")
s5.chainA <- trim.pdb(s5, chain="A", elety="CA")
s6.chainA <- trim.pdb(s6, chain="A", elety="CA")

# This gathers the various B-factor values to be plotted.
s4.b <- s4.chainA$atom$b
s5.b <- s5.chainA$atom$b
s6.b <- s6.chainA$atom$b
```


### 2) Now that I have additional objects to use, let's create the function. We want it to allow any number or set of inputs to be used with our dendrogram.


```{r message=FALSE, warning=FALSE}
# Let's create a function to generalize the process. 

# Here, `...` is used to denote an arbitrary number of inputs.
hcFunction <- function(...){
  
  # Let's paste the original code, but with the `...` to allow for object variation.
  hcAdvanced <- hclust(dist(rbind(...)))
  
  # And now, to plot.
  plot(hcAdvanced)
}
```


### 3) We have now defined `hcFunction()`. Let's attempt to create a dendrogram with all 6 protein chains.


```{r message=FALSE, warning=FALSE}
# Considering that `hcFunction()` was defined with `...` representing the arguments...
hcFunction(s1.b, s2.b, s3.b, s4.b, s5.b, s6.b)
# ...we can add whatever we would like to be in our dendrogram.
```


This code would also work with any combination of proteins, so long as there are at least 3 object inputs in `hcFunction()`.

